---
title: "Impact of Severe Weather Events on Public Health and Economy in the US"
author: "Lisa Yuan"
date: "February 25, 2016"
output: pdf_document
---

## Synonpsis

The purpose of this report is to show the harmful impact of severe weather events on both public
health and the economy. The analysis is performed based on the storm database collected
from the U.S. National Oceanic and Atmospheric Administration (NOAA) from 1950-2011. The
report identifies the weather event types that has the most significant impact on population
health (as measured by fatalities and injuries) and on the economy (as measured by total
damage to property and crops combined in current US dollars). 

Based an my analysis, I find that the severe weather events associated with wind, storm, heat
and drought cause the most damage to public health, whereas severe weather events associated
with flood cause the most damage to the economy.

## Working Directory and Libraries

```{r,messages=FALSE}
setwd("C:/Users/Lisa/version-control/RepData_PeerAssessment2")
library(R.utils)
library(ggplot2)
library(dplyr)
library(grid)
library(gridExtra)

```

## Data Processing 

First, I created a folder for the storm data to be stored.  Then, I downloaded and unzipped the file. 

```{r}
## Create the directory data where the storm data will be downloaded into
if(!file.exists("./data")){
        dir.create("./data")
}

# Download file
if (!"stormData.csv.bz2" %in% dir("./data/")) {
        print ("lets download file")
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                      destfile="data/stormData.csv.bz2")
}

#unzip bz2 file to csv file
if(!file.exists ("./data/StormData.csv")) {
        filePath <- "./data/StormData.csv.bz2"
        destPath <- "./data/StormData.csv"
        bunzip2(filePath, destPath, overwrite=TRUE, remove=FALSE)
}

```

Then, I read the csv file into R environment.

```{r}

if(!"stormData" %in% ls()) {
stormData <- read.csv("./Data/StormData.csv", header = TRUE, sep = ",")
}

```

After the stormData is created as a dataframe, we're able to examine the data. It contains 902,297 rows and
38 columns. The dataset covers the times span from 1950 to 2011. With a quick histogram, it is clearly
evident that the earlier years in the dataset have fewer recorded events, and starting 1994, we see
significant increase of the amount of data available.

```{r}
# Convert the BGN_DATE to year

stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))

# Visulization with histogram

hist(stormData$year, breaks=30)
```

Based on the information from the above histogram and the purpose of this research, I subset data 
keeping only useful columns and rows with year starting at 1990.

```{r}
#Reduce the datasize, only keep the necessary columes
reduced_stormData<- select(stormData, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, 
                           PROPDMGEXP, CROPDMG, CROPDMGEXP, year)
```

By examing the EVTYPE, I have found that the event types can be consolidated into 12 catagories.  These catagories are: 

* Precipitation and Fog
* Dust
* Fire and Volcano
* Flood
* Heat and Drought
* Landslide and Erosion
* Ocean Weather
* Precipitation and Fog
* Thunderstorm and Lightning
* Wind and Storm
* Wintery Weather
* Other


```{r}
#Consolidate the EVTTYPE and created a new variable called EVTCatagory

reduced_stormData$EVTCatagory <- NA
reduced_stormData[grepl("precipitation|rain|hail|drizzle|wet|percip|burst|depression|
                        fog|wall cloud", reduced_stormData$EVTYPE, ignore.case=TRUE), 
                  "EVTCatagory"] <- "Precipitation and Fog"
reduced_stormData[grepl("wind|storm|wnd|tornado|waterspout", reduced_stormData$EVTYPE, 
                        ignore.case = TRUE), "EVTCatagory"] <- "Wind & Storm"
reduced_stormData[grepl("Slide|erosion|slump", reduced_stormData$EVTYPE, ignore.case=TRUE),
                  "EVTCatagory"] <- "Landslide and Erosion"
reduced_stormData[grepl("warmth|warm|heat|dry|hot|drought|thermia|temperature record|
                        record temperature|record high", reduced_stormData$EVTYPE, 
                        ignore.case = TRUE), "EVTCatagory"] <- "Heat and Drought"
reduced_stormData[grepl("cold|chill|ice|Wind Chill|Frost|Freeze|Winter Weater|Winter|
                       Blizzard|Snow|Avalanche", reduced_stormData$EVTYPE, ignore.case=TRUE), 
                  "EVTCatagory"] <- "Wintery Weather"
reduced_stormData[grepl("Tide|Marine|Current|Tsunami|wave|seas|surf|typhoon|hurricane", reduced_stormData$EVTYPE, 
                       ignore.case=TRUE), "EVTCatagory"] <- "Ocean Weather"
reduced_stormData[grepl("flood|blow-out|swells|fld|dam break", reduced_stormData$EVTYPE, 
                        ignore.case=TRUE), "EVTCatagory"] <- "Flood"
reduced_stormData[grepl("dust storm|dust devil", reduced_stormData$EVTYPE, ignore.case=TRUE), 
                  "EVTCatagory"] <- "Dust"
reduced_stormData[grepl("fire|volcanic", reduced_stormData$EVTYPE, ignore.case=TRUE), 
                  "EVTCatagory"] <- "Fire and Volcano"
reduced_stormData[grepl("Thunderstorm|lightning", reduced_stormData$EVTYPE, ignore.case=TRUE), 
                  "EVTCatagory"] <- "Thunderstorm and Lightning"
reduced_stormData$EVTCatagory[is.na(reduced_stormData$EVTCatagory)] <- "Other"
```

I cast the EVTCatagory to a factor variable for later ploting and removed the EVTYPE column. 

```{r}
reduced_stormData$EVTCatagory <- as.factor(reduced_stormData$EVTCatagory)
reduced_stormData$EVTYPE <- NULL
```

In order to assess economic damage casued by severe weather, I calculated total amount using helper functions to convert PROPDMG, PROPDMGEXP into PropDamage, and CROPDMG, CROPDMGEXP into CropDamage

```{r}
# Calculate the damage with conversion

expConvert <- function(x) {
        if(is.numeric(x)) {
                x <- x
        }
        else if (grepl("h", x, ignore.case=TRUE)) {
                x <- 2
        }
        else if (grepl("k", x, ignore.case=TRUE)) {
                x <- 3
        }
        else if (grepl("m", x, ignore.case=TRUE)) {
                x <- 6
        }
        else if (grepl("b", x, ignore.case=TRUE)) {
                x <- 9
        }
        else if(x==""||x==" ") {
                x <- 0
        }
        else{
                x <- NA
        }
        x
}


calamt <- function(num, exp) {
        pow<-expConvert(exp)
        if(is.numeric(num)){
          num <- num * (10^pow)
        }
        
        if(!is.numeric(num)) {
                num <- 0
        }
        
        num
        
}
        
# Getting the absolute value for PropDamage and CropDamage, then add them together to 
# create total damage, 3 new columns are created. 

reduced_stormData$propDamage <- mapply(calamt, reduced_stormData$PROPDMG, reduced_stormData$PROPDMGEXP)
reduced_stormData$cropDamage <- mapply(calamt, reduced_stormData$CROPDMG, reduced_stormData$CROPDMGEXP)
reduced_stormData$totalDamage <- reduced_stormData$propDamage + reduced_stormData$cropDamage
```

I removed the columns no longer needed and subset the data so that I only look at the data starting from 1990.

```{r}
# Remove columns not needed to reduce dataframe size
reduced_stormData$PROPDMG <- NULL
reduced_stormData$PROPDMGEXP <- NULL
reduced_stormData$CROPDMG <- NULL
reduced_stormData$CROPDMGEXP <- NULL

#Furthur reduce the data set by only track the datapoints after 1990

recent_stormData<- filter(reduced_stormData, year >= 1990)
```

## Results 

In this section, I aggregated the total number of fatilities and injuries caused by different types of severe weather catagoy. 

```{r}
#Aggregate the data by event and sorted by fatility/injuries in the descending order

fatality <- aggregate(FATALITIES ~ EVTCatagory, data=recent_stormData, FUN=sum)
fatality <- fatality %>% arrange(desc(FATALITIES))
injuries <- aggregate(INJURIES ~ EVTCatagory, data=recent_stormData, FUN=sum)
injuries <- injuries %>% arrange(desc(INJURIES))
```


```{r}
#Merge fatality and injuries into one dataset
humanDamage <- merge(fatality, injuries, by="EVTCatagory")
```

### Impact on Public Health

Now, let's examine the severe weather damage to public health caused by different events.

```{r}
humanDamage
```

Below are plots for injuries and fatalities impacted by major weather events ordered by severity. 

```{r}
#Fatility plot
ggplot(data=humanDamage, aes(y=FATALITIES, x=reorder(EVTCatagory, -FATALITIES))) + 
        geom_bar(stat="identity", fill="red") +
        scale_y_continuous("Number of Fatalities") +
        ggtitle("Total Fatalities by Severe Weather in US from 1990-2011") + 
        xlab("Weather Event Type") + 
        theme(axis.text.x=element_text(angle=90, hjust=1), text=element_text(size=10))

#Injury plot                                                                  
ggplot(data=humanDamage, aes(y=INJURIES, x=reorder(EVTCatagory,-INJURIES))) + 
        geom_bar(stat="identity", fill="orange") + 
        scale_y_continuous("Number of Injuries") +
        ggtitle("Total Injuries by Severe Weather in US from 1990-2011") +
        xlab("Weather Event Type") + 
        theme(axis.text.x=element_text(angle=90, hjust=1), text=element_text(size=10))
```

Based on the above two histogrms, we find that "Wind and Storm" caused most human injuries, and "Heat and Drought" along with "Wind and Storm" contributed to most human fatalities in the US from 1990 to 2011. 

### Impact on Economy

Now, let's examine what types of severe weather caused most damages to economy. 

```{r}
# Aggregate data for total damage
EconDamage <- aggregate(totalDamage~EVTCatagory, data=recent_stormData, FUN=sum)
EconDamage <- EconDamage %>% arrange(desc(totalDamage))
EconDamage
```

Below is the histogram of total property damage combined with total crop damage affected by severe weather events. 

```{r}
ggplot(data=EconDamage, aes(x=reorder(EVTCatagory, -totalDamage), y=(totalDamage/10^9))) + 
        geom_bar(stat="identity", fill="seagreen4") + 
        scale_y_continuous("Total Economic Damage in US Dollars (billion)") + 
        xlab("Severe Weater Event Type") + 
        ggtitle("Total Economic Damage by Severe Weather in US from 1990-2011")+
        theme(axis.text.x=element_text(angle=90, hjust=1), text=element_text(size=10))
```

Based on the above histogrm, we can conclude that the "Flood" along wtih "Wind & Storm" and "Ocean Weather" caused the most damage to economy in the United States from 1990 to 2011. 

## Conclusion

Based on the data collected from the NOAA and my analysis, we can conclude that the most damaging severe weather types are "Wind and Storm" and "Excessive heat and Drought" to public health.  Whereas the "Flood",  "Wind and Storm" and "Ocean Weather" has the most damaging economic impact. 
